I've been using the `snowfall` package for a while to enable parallel processing in `R` on my Windows machine, where I 
have an 8 core processor. I discovered today that the function `sfLapply` will only work with with an object that has a 
class of `list`. This is really important, because there are many things that are "list-like", and are actually lists at 
heart, but `sfLapply` probably won't like it.

Lets whip up an example. 

```{r exampleStuff, warning=FALSE, message=FALSE}
require(snowfall)
require(GenomicRanges)

gr <- GRanges(seqnames = Rle(c("chr1", "chr2", "chr1", "chr3"), c(1, 3, 2, 4)),
            ranges = IRanges(1:10, width = 10:1, names = head(letters,10)),
            strand = Rle(strand(c("-", "+", "*", "+", "-")), c(1, 2, 2, 3, 2)),
            score = 1:10,
            GC = seq(1, 0, length=10))

gr

class(gr)

grList <- split(gr, seqnames(gr))
grList
class(grList)
```

So we have the `GRanges` object `gr`, and a `GRangesList` in `grList`. Now lets try to do some parallel execution of
finding overlaps of itself.

This is the function we will use in parallel:

```{r functionDef}
returnOverlaps <- function(inObj1, inObj2){
	findOverlaps(inObj1, inObj2, type="any")
}
```

```{r parFail}
sfInit(parallel=T, cpus=2)
sfLibrary(GenomicRanges)

overlap <- sfLapply(grList, returnOverlaps, gr)

sfStop()
```

Ok, we get the error `no method for coercing this S4 class to a vector`. Seems kind of cryptic, at least it did to me.
What about using normal `lapply`?

```{r trylapply}
overlap <- lapply(grList, returnOverlaps, gr)

overlap
```

This works without any errors. Odd. It was only when I was trying to get this to work using `llply` from the `plyr` 
package that I saw a message about `as.default.list` or something like that. So maybe we have to convert the `grList` to 
a good and proper `list` first?

```{r parWork}
sfInit(parallel=T, cpus=2)
sfLibrary(GenomicRanges)

grList <- as.list(grList)
overlap2 <- sfLapply(grList, returnOverlaps, gr)

sfStop()

overlap2
```

It works! I'm not exactly sure why there is this difference, but I thought perhaps I can save someone else a few hours
time figuring it out.

SessionInfo:

```{r sessionInfo, results='asis'}
sessionInfo()
```

Posted on [Blogger](http://robertmflight.blogspot.com/2012/10/sflapply-vs-lapply-in-r.html). [Rmd](https://github.com/rmflight/blogPosts/blob/master/sflapplyVlapply.Rmd), [md](https://github.com/rmflight/blogPosts/blob/master/sflapplyVlapply.md)